Proposal outline/notes

### 2012-06-12 docs/Jesse/Proposal/outline.tex
finished SUPG section minus a reference

We will introduce the SUPG method at an abstract level for illustrative purposes only.  Further details and perspectives on the SUPG method can be found in \cite{SUPG}, as well as in an upcoming book by Hughes.  The convection-diffusion equation can be written as follows:
\[
Lu = \left(L_{\rm adv} + L_{\rm diff} \right) u = f,
\]
where $L_{\rm adv} = \div \left(\beta u\right)$ is the first order advective operator, and $L_{\rm diff} = \epsilon \Delta u$ is the second-order diffusive operator.  Let us assume $u$ to be a linear combination of piecewise-linear basis functions $\phi_i, i = 0,\ldots,N$.  The SUPG method is then to solve
\[
\int_\Omega L u v + \int_\Omega \tau \left(L_{\rm adv}v\right)\left( L u - f\right) = \int_\Omega f v 
\]
for where $\tau$ is the SUPG parameter.  For uniform meshes in 1D, $\tau$ is chosen such that, for $f=0$, the matrix system resulting from SUPG is exactly equal to the finite difference system under ``exact" artificial diffusion.  However, unlike the artificial diffusion method, for $f\neq 0$, the SUPG method still delivers optimal stabilization.  In fact, the SUPG finite element solution in 1D is nothing less than the nodal interpolant and the best $H^1$ approximation of the exact solution, as seen in Figure~\ref{fig:SUPG}.

@@

The idea of emphasizing the upwind portion of a test function is an old idea, introduced in 1977 by Zienkiewicz et al.\ in \cite{zienkUpwind}.  However, the precise amount of upwinding,\footnote{Insufficient upwinding results in a method which still exhibits oscillations and instabilities, while excessive upwinding leads to an overly diffusive method.} as well as the connection to residual-based stabilization methods, were novel to SUPG.  To date, the SUPG is still the most popular stabilization method of choice for convection-diffusion type problems, in both academic and industry applications.

@@

Discontinuous Galerkin (DG) methods form a subclass of FEM; first introduced by Reed and Hill in \cite{Reed:73}, these methods were later analyzed by Cockburn and Shu \cite{CockburnShu:DG} and have rapidly gained popularity for CFD problems. Rather than having a continuous basis where the basis function support spans several element cells, DG opts instead for a discontinuous, piecewise polynomial basis, where, like FV schemes, a numerical flux facilitates communication between discontinuities (unlike FV methods, however, there is no need for a reconstruction step). Advantages of DG methods include easily modified local orders of approximation, easy adaptivity in both $h$ and $p$, and efficient parallelizability.

An additional reason for the popularity of DG methods is that they can be interpreted as stabilized FE methods through appropriate choices of the numerical flux. For example, for the pure convection problem, the solution has only regularity in the streamline direction, but is only $L^2$ in the crosswind direction. As a consequence, the boundary trace of the solution is defined only in the direction of convection. The upwind DG method addresses the above issue by choosing the numerical flux to be the upwind flux; we will illustrate this with a 1D example.

@@

In this case, the DG numerical flux can be viewed as imparting additional regularity than is present in the solution \cite{DPG1,DPG3}.

>>>>

We will introduce the SUPG method at an abstract level for illustrative purposes only.  Further details and perspectives on the SUPG method can be found in \cite{SUPG}, as well as in an upcoming book by Hughes.  The convection-diffusion equation can be written as follows:
\[
Lu = \left(L_{\rm adv} + L_{\rm diff} \right) u = f,
\]
where $L_{\rm adv}u \coloneqq \div \left(\beta u\right)$ is the first order advective operator, and $L_{\rm diff}u \coloneqq \epsilon \Delta u$ is the second-order diffusive operator (then, roughly speaking, $L_{\rm diff} u = 0$).  Let us assume $u$ to be a linear combination of piecewise-linear basis functions $\phi_i, i = 0,\ldots,N$.  The SUPG method is then to solve
\[
\int_\Omega L u v + \int_\Omega \tau \left(L_{\rm adv}v\right)\left( L u - f\right) = \int_\Omega f v 
\]
for where $\tau$ is the SUPG parameter.  For uniform meshes in 1D, $\tau$ is chosen such that, for $f=0$, the matrix system resulting from SUPG is exactly equal to the finite difference system under ``exact" artificial diffusion.  However, unlike the artificial diffusion method, for $f\neq 0$, the SUPG method still delivers optimal stabilization.  In fact, the SUPG finite element solution in 1D is nothing less than the nodal interpolant and the best $H^1$ approximation of the exact solution, as seen in Figure~\ref{fig:SUPG}.

@@

The idea of emphasizing the upwind portion of a test function is an old idea, introduced in 1977 by Zienkiewicz et al.\ in \cite{zienkUpwind}.  However, the precise amount of upwinding,\footnote{Insufficient upwinding results in a method which still exhibits oscillations and instabilities, while excessive upwinding leads to an overly diffusive method.} as well as the connection to residual-based stabilization methods, were novel to SUPG.

The method can be made to ``work" for $p>1$, but the results are not as strong as for linear elements, and there is no clear generalization of SUPG to higher order elements.  Additionally, the stabilization properties of SUPG are weaker in higher dimensions, and the SUPG solution is no longer the $H^1$ best approximation for 2D and 3D problems \todo{find citation}.  Despite these caveats, SUPG is still the most popular stabilization method of choice for convection-diffusion type problems, in both academic and industry applications.

@@

Discontinuous Galerkin (DG) methods form a subclass of FEM; first introduced by Reed and Hill in \cite{Reed:73}, these methods were later analyzed by Cockburn and Shu \cite{CockburnShu:DG} and have rapidly gained popularity for CFD problems. Rather than having a continuous basis where the basis function support spans several element cells, DG opts instead for a discontinuous, piecewise polynomial basis, where, like FV schemes, a numerical flux facilitates communication between discontinuities (unlike FV methods, however, there is no need for a reconstruction step). Advantages of DG methods include the local conservation property, easily modified local orders of approximation, easy adaptivity in both $h$ and $p$, and efficient parallelizability.

An additional reason for the popularity of DG methods is that they can be interpreted as stabilized FE methods through appropriate choices of the numerical flux \todo{add Brezzi citation}. We will illustrate this with the pure convection equation in 1D.

Local conservation: in 1D, numerical solution at outflow is exact.

@@

For example, for the pure convection problem, the solution has only regularity in the streamline direction, but is only $L^2$ in the crosswind direction. As a consequence, the boundary trace of the solution is defined only in the direction of convection. The upwind DG method addresses the above issue by choosing the numerical flux to be the upwind flux; in this case, the DG numerical flux can be viewed as imparting additional regularity than is present in the solution \cite{DPG1,DPG3}.
